\bf Adaption of the Chumbley Score to matching of bullet striation marks

### 2018-02-19 doc/paper.tex
pdf recompiled
meeting monday morning

Compairing pairs of toolmarks with the intention of matching it to a
tool has been studied many times in the past. Extensive examples in
literature for tools and toolmark research ranging from screwdrivers to
groove pliers to slip-joint pliers can be found in the work of
\citet{manytoolmarks1}, \citet{manytoolmarks2}, \citet{miller},
\cite{afte-chumbley} and many more. In comparison to this, same source
matching of bullets to firearms has not been examined as prominently as
that of toolmarks, with even less information available on validity of
methods and error rates associated with firearms examination. The
National Academy of Sciences in its report in 2009 \citep{NAS:2009}
discussed the need for determining error rates in methods proposed for
firearms examination. In the context of same source matching and error
rates determination, as seen in the case of most forensic applications,
the first step involves identification of unique features that are
characteristic of the object at hand. For the case of bullets and
firearms, striation marks on the surface of the bullet are considered to
be such markings that can be used in methods for same source matching.
These marks are often a product of rifling and impurities and defects
due to manufacturing in the barrel of the gun, which leads to engravings
on the bullet surface \citep{afte-article1992}. In current practice,
firearm examiners invariably make visual comparisons of bullet striae
and use visual assesment tools to dignify bullets as being matches and
non-matches. One way of accomplishing anykind of comparison between
bullets is to do a comparison between surface marking of two or more
bullet lands. Bullet Lands are considered to be areas between grooves
made by the rifling action of the barrel. These marking are considered
to be unique. The land engraved markings or sometimes termed as Bullet
profiles \citep{aoas} are striation marks made on Bullet lands and often
used for these land to land comparisons. Bullet Signatures is another
word used in literature as seen in the work of \citet{chu2013} and
\citet{aoas}. In our context bullet signatures refer to a processed
version of the raw land engraved markings or profiles. The generation of
bullet signatures involves first extraction of a bullet profile by
taking the cross-sectional of the surface at a given height and then
using loess fits to model the structure. The residuals of this fit are
called signatures, which are considered to be noise free and a good
reflection of the class charecteristics and unique features of a bullet.
A more detailed version of the extraction technique of signatures is
discussed by \citet{aoas}, where comprehensive details about the height
at which profile is to be selected, removing curvature, smoothing,
identifying groove locations are explained.

In the study conducted by \citet{aoas} a machine learning based
algorithm was developed for same source matching of bullets and error
rates were discussed using the database from the Hamby Study
\citep{hamby}. In this paper, we first try to adapt a deterministic
algorithm and method developed for toolmarks by \citet{chumbley} and
improved by \citet{hadler}, to bullets. Then we consequently discuss
about the efforts in doing so along with the associated error rates. The
data used in this paper also belongs to the Hamby Study \citep{hamby}.
This gives us a common platform for comparing the performance of the
chumbley method on bullets with an already existing method proposed by
\citet{aoas} for bullets. The proposed algorithm and method of
\citet{chumbley}, in their paper, compares two toolmarks with the
intention of determining if it comes from the same source (same tool).
The method also provides a means to determine error rates and claims to
reduce subject bias. As mentioned earlier, subject bias and error rate
determination have been a long standing issue in firearm examination
\citep{NAS:2009}. This remains one of the motivations to explore the
adaptibility of the Chumbley score methodology to bullets.
\citet{chumbley} used an empirical based setup to validate their
proposed algorithm and quantitative method which calculates a U-statisic
for the purpose of classification of toolmarks as matching or
non-matching. The data for their study was obtained from 50 sequentially
manufactured screwdriver tips, and preselected comparison window sizes
were given as inputs to the algorithm. The algorithm then compares the
two toolmarks and comes up with a U-statistic and an associated p-value
to designate them as matches or non-matches. The performance for every
100 comparisons, of the algorithm proposed by \citet{chumbley} and the
improvement proposed by \citet{hadler} are listed in the table below.

@@

Bullets are much smaller in length, width, are not flat and curved in
the cross-sectional topography as opposed to tools like screw driver
tips which produced longer and pronounced markings. This means the
makings made by barrels in comparison to toolmarks may have a problem in
distinctiveness. The majority of Bullet profiles and signatures
extracted by procedures mentioned by \citet{aoas} are almost 1/4 th the
size of toolmarks as used by \citet{chumbley} or even smaller.
Striations on Bullets are made on their curved surfaces, whereas the
algorithm developed by \citet{chumbley} and \citet{hadler} has only been
tested for flatter and wider surfaces which have negligible curvature.
Therefore, using methods proposed for toolmarks may need adaptation in
order to give tangible results for bullets. Moreover, in order to to get
flat bullet signatures and remove the curvatures some kind of smoothing
needs to be applied as a pre-step which needs further investigation as
to whether the level of smoothing does effect the working of the
algorithm on Bullets.

Also when in the optimization step, the Window of optimization for
bullets will be shorter as the signatures are smaller. The idea is to
keep the number of windows of optimization sufficiently large, which
means shorter Trace segments (partition of signature or toolmark with
length = size of window of optimization) that lets us compare smaller
segments of one signature to another. This introduces a problem as, if
we go too small in the window of optimization, the unique features of
the trace segments are lost and seem similar, while too large sizes
vastly reduces the weight of small features that would otherwise
uniquely classify a signature and hence identify the region of
agreement.

Thus the Window of Optimization has a direct influence on whether we are
Falsely rejecting the null when it is true (Type I) or Falsely accepting
the null when it is false (Type II) making identification of the optimum
size of window of optimization very important. This raises important
questions about what parameter settings need to be chosen for bullet
land comparison. Something similar is done in \citet{afte-chumbley} for
toolmark comparisons of slip-joint pliers where optimum window sizes are
determined. There is a need to figure out the best parameter settings
which minimizes the errors for unsmoothed markings and pre-processed
signatures as determined by \citet{aoas}. An analysis of these error
rates and comparison with other methods will help us understand the
adaptibility of the chumbley score to bullets.

\hh{figure \ref{fig:rgl} needs to be mentioned somewhere in the write-up}

@@

\section{The Chumbley Score Test}\label{the-chumbley-score-test}

The Chumbley score algorithm takes input as two vectorized processes
\(x(t_1)\), \(t_1 = 1,2,...T_1\) and \(y(t_2)\), \(t_2 = 1,2...T_2\)
which denote two sets of marks or striae. Here the processes are of the
form \(z(s)\) which is a spatial process for some \(t\), while
\(z(s_1)\) is the realization of the process in \(s_1\). The \(t_1\) and
\(t_2\) can be better understood as equally spaced pixel locations for
the two marks under consideration, where a `pixel' is the resolution of
the confocal light microscope. In the case for bullet signatures a pixel
corresponds to 0.645 microns. These marks or striae are potentially from
two different bullets or two different toolmarks whose source needs to
be identified as being same or different. The marks or striae are
indexed by the pixel location where \(t_1\) is for the first striae
referred to as \(x\) and \(t_2\) is for the second striae which is
referred as \(y\). \(T_1\) and \(T_2\) being the respective lengths of
the markings, need not be the same but are usually of similar lenghtsThe
similarity is then judged by the algorithm on the basis of
cross-correlation of a fixed and constant number consecutive pixels say
\(k\) taken from the two indexed marks \(x(t_1)\) and \(y(t_2)\) such
that in theory \(k\) remains smaller that length of the two striae or
marks which is the same as \(t_1\) and \(t_2\). Depending on the what
stage of the algorithm we are in, matching of different pixel lengths
and locations is done, which at the end effectively compares all
possible windows that would guarantee in quantifying the two marks or
striae as coming from the same source or not, which also lets us assess
the error rates by checking for a large number of cases.

The algorithm works in two phases, namely, an optimization step and a
validation step, at the end of which a Mann Whitney U statistic is
calculated. A pre-processing step to the algorithm is to choose a
coarseness value which is used as a parameter to the LOWESS smoothing
function. The coarseness essentially gives the proportion of points
which influence the smooth at each value, which means larger values lead
to more smoothness. The LOWESS smoothing is applied to each of two sets
of vectorized striae or marks \(x(t_1)\) and \(y(t_2)\), before
proceeding to the algorithm.

\citet{hadler} in their paper proposed an improvement to this algorithm
by trying to remove mutual dependence of parameters (due to serial
correlation in surface depth values of a toolmark and because of a
random sampling sub step in the validation phase which makes a group of
pixels to be chosen more than once and hence introduces lack of
independence) in certain steps, especially because the Mann-Whitney U
statistic that is later calculated in the algorithm and used as a
measure to differentiate between matches and non-matches works under the
assumption of independence of parameters.

Since we are only interested in a non-parameteric U statistic, Hadler et
al. proposed a normalization procedure in the Validation step that goes
to some extent to address the issue of mutual dependence. Also the same
shift and different shift substep was modified to use a deterministic
rule for sampling sam-shhift and dofferent shift-samples as opposed to
the originally proposed random samples.

The data that is used by \citet{chumbley} is generated by a surface
profilometer that gives the height in terms of distance along a linear
trace. This is taken perpendicular to the striations present in the
toolmark. Two such trace are then compared using the algorithm.

\subsection{Detailed algorithm}\label{detailed-algorithm}

\subsubsection{Optimization step}\label{optimization-step}

The idea behind this step is to first identify the area of best
agreement in the two toolmark data. Comparison window size is predefined
by the user, which in case of screwdriver toolmarks was chosen by
Chumbley et al and Hadler et al as around 10 percent of the length of
the toolmarks, which was around 500. This window is henceforth referred
to as Window of optimization.

A maximum correlation statistic is used to identify the region of best
agreement, with the maximum usually seen being near 1 for both cases
which is what we intuitively expect for matches, and something which we
do not intuitively expect for non matches

First, there are a very large number of cross-correlations calculated
for the two series of striae or marks \(x(t_1)\) and \(y(t_2)\), for eg
if the window of optimization was defined as 200 and toolmark pixel
length is around 1000 then we have 1200-(200-1) = 1001.

This is the number of windows we have for one toolmark and each window
is compared with all windows of the second toolmark (the number of
windows is again similar), and the window with the maximum correlation
is identified to be the region where the toolmarks are in maximum
agreement with each other

\subsubsection{Validation Step}\label{validation-step}

\paragraph{Same-shift:}\label{same-shift}

In this step a series of windows are chosen at random (originally by
chumbley et al) and deterministically (by hadler et al), but at a common
distance (rigid-shift) from the window identified as the region of best
agreement in the vaildation step. The correlation of these windows would
be as intuitively assumed i.e.~lower than the maximum correlation window
(Optimization step), but the significance is that these same shift
windows will still have large enough correlation values for two
toolmarks or signatures that are in reality a match.

This also validates that if in the optimization step the maximum
correlation window chosen (with correlation value near 1) was by
accident (like in case of signatures that in reality are a non-match),
all these same- shift correlations (A fixed number of these trace
segments are identified) would not be anywhere large enough for all
same-shift windows.

@@

\paragraph{Different Shift:}\label{different-shift}

Primary reason for this substep is to give perspective to the
correlation values of the same shift window correlation values.

This time there are no rigid-shifts but different shifts (distance from
Window of Opt with max correlation) chosen randomly by \citet{chumbley}
and deterministically by Hadler et al. such that there is an equal
possibilty of comparing a trace segment from one signature or toolmark
to any one in the second signature or toolamrk.

Neither of above sets of correlation are allowed to include the maximum
correlation window as identified earlier.

Therefore the assumption is that if two toolmarks or signatures match
each other the same-shift correlations would be larger than the
different-shift windows, and if they are not a match the correlations in
the two sets will be very similar.

\subsubsection{U Statistic:}\label{u-statistic}

This is computed from the joint rank of all correlations of both the
same and different shift samples. As given by \citet{hadler}

Null Hypothesis: If the toolmarks were not match i.e not made by the
same tool.

Let ns and nd be the number of same shift and different shift windows
\[N = n_{s} + n_{d}\]

The mann whitney U statistic is given by
\[U =\sum^{ns}_{i=1}R_{s}\left( i\right)\]

with the standardized version which includes provision for rank ties

\[\overline{U}= \dfrac{U-M}{\sqrt{V}}\]

where prior to normalization the U-statistic has the mean as

\[M = n_{s}\left(\dfrac{N+1}{2}\right)\]

and variance

\[V = \dfrac{n_{s}n_{d}}{N\left(N-1\right)}\left[\Sigma^{n_{s}}R_{s}\left(i\right)^{2}+\Sigma^{n_{d}}R_{d}\left(j\right)^{2}\right] -\dfrac{n_{s} n_{d}\left(N+1\right)^{2}}{4\left(N-1\right)}\]

\subsection{Testing setup}\label{testing-setup}

@@

\centering
\resizebox{\linewidth}{!}{\resizebox{\linewidth}{!}{\begin{tabular}[t]{lrrrrrrrrrrrrrrrrrr}
\toprule
wo & 50 & 50 & 60 & 60 & 80 & 80 & 90 & 90 & 100 & 100 & 110 & 110 & 120 & 120 & 120 & 120 & 120 & 120\\
wv & 30 & 50 & 30 & 50 & 30 & 50 & 30 & 50 & 30 & 50 & 30 & 50 & 10 & 20 & 30 & 40 & 50 & 60\\
\bottomrule
\end{tabular}}}
\centering
\begin{tabular}[t]{lrrrrrrrrrrrrrrr}
\toprule
wo & 130 & 130 & 140 & 140 & 150 & 150 & 160 & 160 & 200 & 200 & 200 & 240 & 240 & 280 & 280\\
wv & 30 & 50 & 30 & 50 & 30 & 50 & 30 & 50 & 20 & 30 & 50 & 30 & 50 & 30 & 50\\
\bottomrule
\end{tabular}
\end{table}

@@

\centering
\begin{tabular}[t]{llrl}
\toprule
\hspace{1em}\hspace{1em}signif & match & Freq & Type\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Size of Optimization Window = 280}}\\
\hspace{1em}FALSE & FALSE & 78909 & True Negative\\
\hspace{1em}TRUE & FALSE & 4192 & False Positive (Type I)\\
\hspace{1em}FALSE & TRUE & 446 & False Negative (Type II)\\
\hspace{1em}TRUE & TRUE & 731 & True Positive\\
\bottomrule
\end{tabular}
\centering
\begin{tabular}[t]{llrl}
\toprule
signif & match & Freq & Type\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Size of Optimization Window = 120}}\\
\hspace{1em}FALSE & FALSE & 79249 & True Negative\\
\hspace{1em}TRUE & FALSE & 4523 & False Positive (Type I)\\
\hspace{1em}FALSE & TRUE & 386 & False Negative (Type II)\\
\hspace{1em}TRUE & TRUE & 854 & True Positive\\
\bottomrule
\end{tabular}
\centering
\begin{tabular}[t]{llrl}
\toprule
signif & match & Freq & Type\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Size of Optimization Window = 80}}\\
\hspace{1em}FALSE & FALSE & 79477 & True Negative\\
\hspace{1em}TRUE & FALSE & 4503 & False Positive (Type I)\\
\hspace{1em}FALSE & TRUE & 463 & False Negative (Type II)\\
\hspace{1em}TRUE & TRUE & 781 & True Positive\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:lms}Estimates of the increase in percent of failed tests corresponding to a 100 point increase in the optimization window.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
match & wv & estimate & std.error\\
\midrule
FALSE & 30 & 0.452 & 0.027\\
FALSE & 50 & 0.922 & 0.035\\
TRUE & 30 & 0.037 & 0.004\\
TRUE & 50 & 0.056 & 0.005\\
\bottomrule
\end{tabular}
\end{table}

@@

{\centering \includegraphics[width=\textwidth]{figures/wo120-profile-1}

\caption{The figure shows the Type II error for different validation window sizes and different chosen nominal alpha levels when the size of the optimization window = 120}\label{fig:wo120-profile}
\end{figure}

>>>>

\hh{XXX first sentence: Same source analyses are a major part of an Forensic Toolmark examiners job. XXX second sentence: describe standard practice: In current practice firearm examiners  make these comparisons visually under a comparison microscope and report XXX what is being reported: AFTE. XXX find: AFTE theory of identification as it relates to toolmarks  }
\hh{One of the main criticisms in this toolmark examination first raised by the National Research Council in 2009 [@NAS:2009] and later emphasized further by the President's Council of Advisors on Science and Technology [@pcast2016] are the amount of subjectivity in the assessment and the lack of error rates. }
\hh{XXX we have digitized images, e.g. profilometers or Confocal microscopy allow us to measure the surface}

\hh{Using a statistical approach removes both subjectivity from the assessment and allows us to provide error rates for both false positive and false negative identifications. }
\hh{Multiple toolmarks have been studied in the literature:  \citep{manytoolmarks1, manytoolmarks2, chumbley} have studied screwdriver marks digitized using a profilometer,  \citep{manytoolmarks2} investigate groove pliers XXX data? XXX  to slip-joint pliers \citep{afte-chumbley} XXX. In all of these cases, the statistical test used, is a variation of the Chumbley scoring method, first introduced by XXX and later improved by XXX. }

\hh{XXX In this pper, we are investigating the applicability of the Chumbley scoring method by @hadler to assessing striation marks on bullets for same-source identification. }

\hh{Striation marks on bullets are made by impurities in the barrel. As the bullet travels through the barrel, these imperfections leave "scratches' on the bullet surface. Typically, only striation marks in the land engraved areas (LEAs) are considered (XXX citation) [@afte-article1992]. Compared to toolmarks made by screwdrivers striation marks on bullets are typically much smaller, both in length and in width. Bullets also have a curved cross-sectional topography. 
}

Compairing pairs of toolmarks with the intention of matching it to a
tool has been studied many times in the past. Extensive examples can be
found in literature for tools and toolmark research ranging from
screwdrivers \citep{manytoolmarks1, manytoolmarks2, chumbley} to groove
pliers \citep{manytoolmarks2} to slip-joint pliers \citep{afte-chumbley}
and many more. In comparison to this, same source matching of bullets to
firearms has not been examined as prominently as that of toolmarks. Even
less information is available on validity of methods and error rates
associated with firearms examination. The National Academy of Sciences
in its report in 2009 \citep{NAS:2009} discussed the need for
determining error rates in methods proposed for firearms examination. As
seen in the case of most forensic applications, the first step for same
source matching and error rate determination involves identification of
unique features that are characteristic of the object at hand. For the
case of bullets and firearms, striation marks on the surface of the
bullet are considered to be such markings that can be used in methods
for same source matching. These marks are often a product of rifling and
impurities and defects due to manufacturing in the barrel of the gun,
which leads to engravings on the bullet surface
\citep{afte-article1992}. In current practice, firearm examiners
invariably make visual comparisons of bullet striae and use visual
assesment tools to dignify bullets as being matches and non-matches. One
way of accomplishing anykind of comparison between bullets is to do a
comparison between surface marking of two or more bullet lands. Bullet
Lands are considered to be areas between grooves made by the rifling
action of the barrel and the markings on them are considered to be
unique. The land engraved markings or sometimes termed as Bullet
profiles \citep{aoas} are striation marks made on Bullet lands and often
used for these land to land comparisons. Bullet Signatures is another
word used in literature as seen in the work of \citet{chu2013} and
\citet{aoas}. In our context bullet signatures refer to a processed
version of the raw land engraved markings or profiles. The generation of
bullet signatures involves first extraction of a bullet profile by
taking the cross-sectional of the surface at a given height and then
using loess fits to model the structure. The residuals of this fit are
called signatures, which are considered to be noise free and a good
reflection of the class charecteristics and unique features of a bullet.
A more detailed version of the extraction technique of signatures is
discussed by \citet{aoas}, where comprehensive details about the height
at which profile is to be selected, removing curvature, smoothing,
identifying groove locations are explained. Figure \ref{fig:rgl} shows
us how the extracted signature from a bullet land lines up with the
image of the land from which it was extracted. We can see also see in
the figure how the depth and relative position of the striation markings
seen in the image are interpreted as the signature.

As mentioned earlier, subject bias and error rate determination have
been a long standing issue in firearm examination \citep{NAS:2009}. This
issue of removing subjectivity and making firearm analysis objective was
also raised in the \citet{pcast2016} report. For an objective analysis,
scientific principle testability and error rate determination are
considered to be defining aspects. Identification of the uncertainities
associated with a method of comparison and coming up with a systemized
method of quantification, is therefore important. Associating
mathematical probabilites in a systematic manner to these methods are
one such way of quantification. In a previous study conducted by
\citet{aoas} a machine learning based algorithm was developed for same
source matching of bullets and error rates were discussed using the
database from the Hamby Study \citep{hamby}. The method proposed by
\citet{chumbley} (and later improved by \citet{hadler}) also provides a
means to determine error rates and claims to reduce subject bias.
Therefore, giving us a strong motivation to explore the adaptibility of
the Chumbley score methodology to bullets. In this paper, we take this
toolmark based methodology and try to apply and adapt it to bullets.
After this, we consequently discuss about the efforts in doing so, along
with the associated error rates. The data used in this paper also
belongs to the Hamby Study \citep{hamby}. This gives us a common
platform for comparing the performance of the chumbley method on bullets
with the already existing method proposed by \citet{aoas} for bullets.
As a brief overview, \citet{chumbley}, in their paper, compare two
toolmarks with the intention of determining if it comes from the same
tool. They use an empirical based setup to validate their proposed
algorithm and quantitative method which calculates a U-statisic for the
purpose of classification of toolmarks as matching or non-matching. The
data for their study was obtained from 50 sequentially manufactured
screwdriver tips, and preselected comparison window sizes were given as
inputs to the algorithm. The algorithm then compares the two toolmarks
and comes up with a Mann-Whitney U-statistic and an associated p-value
to designate them as matches or non-matches. The performance for every
100 comparisons, of the algorithm proposed by \citet{chumbley} and the
improvement proposed by \citet{hadler} are listed in the table below.

@@

Bullet striations are different from toolmarks. Bullet markings are
typically produced by the barrel on the bullets, while toolmarks can be
made by tool tips on different surfaces. Tools like screw driver tips as
used by \citet{chumbley}, produce longer and pronounced markings.
Bullets on the other hand are much smaller in length, width, and curved
in the cross-sectional topography. This means the markings made by
barrels as opposed to markings made by tools, may have a problem in
distinctiveness. The majority of Bullet profiles and signatures of the
Hamby study \citep{hamby} used in this paper are extracted by procedures
mentioned by \citet{aoas}. The Ruger pistol barrels used in the Hamby
study have a 9mm caliber. This means that bullets used in these pistols
have land markings of almost \((1/6)^{th}\) the size of screwdriver
toolmarks used by \citet{chumbley}. Also, striations on bullets are made
on their curved surfaces, whereas the algorithm developed by
\citet{chumbley} and \citet{hadler} has only been tested for flatter and
wider surfaces which have negligible curvature. Therefore, using methods
proposed for toolmarks may need adaptation in order to give tangible
results for bullets. Moreover, in order to get flat bullet signatures
and remove the curvatures some kind of smoothing needs to be applied as
a pre-step wihch needs further investigation as to whether the level of
smoothing does effect the working of the algorithm on Bullets.

The Chumbley method and algorithm works on the premise of comparison
windows which are supposed to represent small segments of pixels of
equal sizes in the two markings that are to be compared in a
predetermined fashion. They term them as windows of optimization and
validation. The optimization window is selected in the first step called
the optimization step where areas/windows of best agreement between the
the two markings are found. The window of optimization for bullet
striations are shorter as bullet signatures are smaller compared to
toolmarks . The idea is to keep the number of windows of optimization
sufficiently large. This means we have shorter trace segments that lets
us compare smaller segments of one signature to another than those for
toolmarks. Here, trace segments are the partitions of marking with the
length of the partition equal to the size of window of optimization.
This introduces a problem as, if we go too small in the window of
optimization, the unique features of the trace segments are lost and
seem similar, while too large sizes vastly reduces the weight of small
features that would otherwise uniquely classify a signature and hence
identify the region of agreement. The problem of distinctiveness in the
structure of bullet striations may also lead to potential failure of the
algorithm to identify correct windows of agreements or not be able to
identify these windows at all. This may lead to incorrect classification
of bullets as matches or non-matches.

Thus, the comaparison windows have a direct influence on the false
positives and false negatives, making identification of the optimum
sizes of these windows very important. This raises important questions
about how and what parameter settings need to be chosen for bullet land
comparison. Something similar is done in \citet{afte-chumbley} for
toolmark comparisons of slip-joint pliers where optimum window sizes are
determined. There is a need to figure out the best parameter settings
which minimizes the errors for unsmoothed markings or profiles and
pre-processed signatures. An analysis of these error rates and
comparison with other methods will help us understand the adaptibility
of the chumbley score to bullets.

@@

\subsection{The Chumbley Score Test}\label{the-chumbley-score-test}

The Chumbley score algorithm takes input as two vectorized processes.
Here the processes are of the form \(z(t)\) which is a spatial process
for some location indexed with \(t\). This means that while \(z(t)\)
represents any striae, \(z(t_1)\) is a realization of the spatial
process. \(t\) here denotes a vector of equally spaced pixel locations
for the striation marks under consideration. The word `pixel' refers to
the resolution of the confocal light microscope. In the case for bullet
signatures extracted from NIST ballistics database \citep{nist} of the
Hamby study, a pixel corresponds to approximately 1.56 microns. The
vectorized processes representing two sets of striaes are therefore
shown as \(x(t_1)\), \(t_1 = 1,2,...T_1\) and \(y(t_2)\),
\(t_2 = 1,2...T_2\). Here \(x\) and \(y\) denote two instances of the
process \(z(*)\) which means two striation marks, while the indexing
\(t\) or pixel range is shown as \(t_1\) and \(t_2\) for \(x\) and \(y\)
respectively. This representation of \(t\) as \(t_1\) and \(t_2\) lets
us identify them as either of same or different lengths. The striation
marks under consideration are potentially from two different bullets
whose source needs to be identified as being same or different. \(T_1\)
and \(T_2\), as represented above, are the final pixel indexes of each
marking and therefore give the respective lengths of the markings. The
similarity in the two markings is then judged by the algorithm on the
basis of cross-correlation of a fixed and constant number consecutive
pixels (say \(k\)) taken from the two markings. This can for example be
\(k\) taken from the two indexed marking \(x(t_1)\) and \(y(t_2)\) such
that in theory \(k\) remains smaller than length of the two striae or
marks. Depending on what stage of the algorithm we are in, matching of
different pixel lengths and locations is done. This in the end,
effectively compares all possible windows that would guarantee in
quantifying the two marks or striae as coming from the same source or
not.

The algorithm works in two phases, namely, an optimization step and a
validation step, at the end of which a Mann Whitney U statistic is
calculated as a measure to differentiate between matches and
non-matches. A pre-processing step to the algorithm is to choose a
coarseness value which is used as a parameter to the lowess smoothing
function. The coarseness essentially gives the proportion of points
which influence the smooth at each value, which means larger values lead
to more smoothness. The lowess smoothing is applied to each of two sets
of vectorized striae or marks \(x(t_1)\) and \(y(t_2)\), before
proceeding to the algorithm. The algorithm starts with the optimization
step where the area of best agreement in the two markings being compared
is identified. Comparison window size is predefined. Each window of the
first marking is compared with all windows of the second marking. A
maximum correlation statistic is used to identify the region of best
agreement, with the maximum usually seen being near 1 for both cases
which is what is intuitively expected for matches, and not expected
intuitively for non matches. \citet{hadler} in their paper proposed an
improvement to this algorithm by trying to remove mutual dependence of
parameters. The mututal dependence was due to the serial correlation in
surface depth values of a marking. Also a random sampling sub step in
the validation phase makes a group of pixels to be chosen more than once
and hence introduces lack of independence in certain steps. The reason
for removing the mutual dependence was the Mann-Whitney U statistic
works under the assumption of independence of parameters. The validation
step builds on this with two sub-steps namely, same shift and different
shift. As the aim is to come up with a non-parameteric U statistic,
\citet{hadler} proposed a normalization procedure in the validation step
which goes to some extent to address the issue of mutual dependence. A
series of windows are chosen in both the same and different shift
sub-steps for the purpose of comparison. Both substeps were modified by
\citet{hadler} who introduced a deterministic rule for sampling of
windows in the same-shift and different shift as opposed to the
originally proposed random sampling by \citet{chumbley}. In th same
shift substep the chosen series of windows are at a common distance
(rigid-shifts) from the window earlier identified as the region of best
agreement in the optimization step. The correlation of these windows
generally turn out to be lower than the maximum correlation window. The
significance is that these same shift windows still have large enough
correlation values for the two markings being compared that are in
reality a match. Any choice of maximum correlation windows (where
correlation values are often near 1) for non-matches in the optimization
step are also validated by this sub-step. This is because the same-shift
correlations would not be anywhere as large for same-shift windows when
all windows in the sub-step are compared. The different shift substep on
the other hand gives perspective to the correlation values of the same
shift window correlation values. There are no rigid-shifts but different
shifts where distance from the maximum correlation window are chosen
randomly by \citet{chumbley} and deterministically as per
\citet{hadler}. This means that there is an equal possibilty of
comparing a trace segment from one marking to any trace segment or
window in the second marking.

Neither of above sets of correlation in the two sub-steps are allowed to
include the maximum correlation window as identified earlier. Therefore
the assumption is that if two markings match each other, the same-shift
correlations would be larger than the different-shift windows. And if
they are not a match the correlations in the two sets will be very
similar. The U-statistic tests for the null hypothesis that the two
markings are not a match and are therefore not made by the same source.
As given by \citet{hadler}, it is computed from the joint rank of all
correlations of both the same and different shift samples.

@@

\section{Simulation setup}\label{simulation-setup}

@@

\centering
\resizebox{\linewidth}{!}{\resizebox{\linewidth}{!}{\begin{tabular}[t]{lrrrrrrrrrrrrrrrrrr}
\toprule
wo & 50 & 50 & 60 & 60 & 80 & 80 & 90 & 90 & 100 & 100 & 110 & 110 & 120 & 120 & 120 & 120 & 120 & 130\\
wv & 30 & 50 & 30 & 50 & 30 & 50 & 30 & 50 & 30 & 50 & 30 & 50 & 10 & 20 & 30 & 50 & 60 & 30\\
\bottomrule
\end{tabular}}}
\centering
\begin{tabular}[t]{lrrrrrrrrrrrrrrr}
\toprule
wo & 130 & 140 & 140 & 150 & 150 & 160 & 160 & 200 & 200 & 200 & 240 & 240 & 280 & 280 & 320\\
wv & 50 & 30 & 50 & 30 & 50 & 30 & 50 & 20 & 30 & 50 & 30 & 50 & 30 & 50 & 30\\
\bottomrule
\end{tabular}
\end{table}

@@

\centering
\begin{tabular}[t]{llrl}
\toprule
\hspace{1em}\hspace{1em}signif & match & Freq & Type\\
\midrule
\addlinespace[0.5em]
\multicolumn{4}{l}{\textbf{Size of Optimization Window = 280}}\\
\hspace{1em}FALSE & FALSE & 78909 & True Negative\\
\hspace{1em}TRUE & FALSE & 4192 & False Positive (Type I)\\
\hspace{1em}FALSE & TRUE & 446 & False Negative (Type II)\\
\hspace{1em}TRUE & TRUE & 731 & True Positive\\
\bottomrule
\end{tabular}
\centering
\begin{tabular}[t]{llrl}
\toprule
signif & match & Freq & Type\\
\midrule
\addlinespace[0.5em]
\multicolumn{4}{l}{\textbf{Size of Optimization Window = 120}}\\
\hspace{1em}FALSE & FALSE & 79249 & True Negative\\
\hspace{1em}TRUE & FALSE & 4523 & False Positive (Type I)\\
\hspace{1em}FALSE & TRUE & 386 & False Negative (Type II)\\
\hspace{1em}TRUE & TRUE & 854 & True Positive\\
\bottomrule
\end{tabular}
\centering
\begin{tabular}[t]{llrl}
\toprule
signif & match & Freq & Type\\
\midrule
\addlinespace[0.5em]
\multicolumn{4}{l}{\textbf{Size of Optimization Window = 80}}\\
\hspace{1em}FALSE & FALSE & 79477 & True Negative\\
\hspace{1em}TRUE & FALSE & 4503 & False Positive (Type I)\\
\hspace{1em}FALSE & TRUE & 463 & False Negative (Type II)\\
\hspace{1em}TRUE & TRUE & 781 & True Positive\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:lms}Estimates of the increase in percent of failed tests corresponding to a 100 point increase in the optimization window.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
match & wv & estimate & std.error\\
\midrule
FALSE & 30 & 0.503 & 0.028\\
FALSE & 50 & 0.922 & 0.035\\
TRUE & 30 & 0.043 & 0.004\\
TRUE & 50 & 0.056 & 0.005\\
\bottomrule
\end{tabular}
\end{table}

@@
